Challenges in Developing Great Quasi-Monte Carlo Software

### 2023-07-12 MCQMC2022ChallengesQMCSoftwareArticle_revision_1.tex
aleksei updates and sign off

%%%%%%% Author sign off on this revised submission, last revision 7/12/2023
% Sou-Cheng Choi signed off on 
% Yuhan Ding signed off on 
% Fred J. Hickernell signed off on 
% Jagadeeswaran Rathinavel signed off on
% Aleksei Sorokin signed off on

@@

\begin{figure}[t]
    \centering
    \raisebox{0.5cm}{\includegraphics[width=0.4\textwidth]{BeamDrawing.png}}
    \includegraphics[width=0.5\textwidth]{cantileveredbeamwords.eps}
    \includegraphics[width=1\textwidth]{ldparallelbeam.eps}
    \caption{The upper left shows the piecewise log effective stiffness (image reproduced from~\cite{ParSee22a} with permission). The expected deflection of a cantilevered beam with random log effective stiffness is shown in the upper right.  QMCPy computes the answer in serial and in parallel using UM-Bridge and docker containers.  For small error tolerances, the time and sample size is $\calO(\varepsilon^{-1})$ (lower right), and parallel processing gives a speed-up factor of four to five (lower left). These plots may be reproduced using the Jupyter notebook \cite{MCQMC2022Figs}.  }
    \label{CDHJSfigbeam}
\end{figure}

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \cite{UMBridge23}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$, and their derivatives if applicable. UM-Bridge allows coupling of UQ codes to models regardless of the programming languages.
{eq:cubSummary}
We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam of length~$L$, with a random log effective stiffness $m(\cdot)$, assumed to be piecewise constant over three regions, each of which with a uniform distribution of uncertainty~\cite{ParSee22a}, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  The beam position $x$ is discretized at $N=31$ equidistant points, and
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. 
%approximates $u(x)$ and $m(x)$ at the nodes, respectively.
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$.
The beam model was written in C++ using the MUQ's model framework~\cite{MUQ2}. It is one of the UM-Bridge's benchmark problem library, available as a container.

@@

Although Python is popular and versatile, but it can be slower than highly optimized implementations in languages like Java or C++.  Developers of QMC software in Python  developers would do well to investigate interfaces to optimized code in other languages or consider other code acceleration techniques provided by libraries such as Numba, JAX, Joblib, or Ray.  With time, languages rise and fall in popularity, and those who wish to provide great QMC software may need to pivot to new software environments.

>>>>

%%%%%%% Author sign off on this revised submission, last revision 7/12/2023
% Sou-Cheng Choi signed off on 
% Yuhan Ding signed off on 
% Fred J. Hickernell signed off on 
% Jagadeeswaran Rathinavel signed off on
% Aleksei Sorokin signed off on 7/12/2023

@@

\begin{figure}[t]
    \centering
    \raisebox{0.5cm}{\includegraphics[width=0.4\textwidth]{BeamDrawing.png}}
    \includegraphics[width=0.5\textwidth]{cantileveredbeamwords.eps}
    \includegraphics[width=1\textwidth]{ldparallelbeam.eps}
    \caption{The upper left plot shows a piecewise log effective stiffness (image reproduced from~\cite{ParSee22a} with permission). The upper right plot shows the expected deflection of a cantilevered beam with random log effective stiffness. In the lower plots, QMCPy computes the answer in serial and in parallel using UM-Bridge and docker containers.  For small error tolerances, the time and sample size are $\calO(\varepsilon^{-1})$ and parallel processing gives a speed-up factor of four to five. These plots may be reproduced using the Jupyter notebook \cite{MCQMC2022Figs}.  }
    \label{CDHJSfigbeam}
\end{figure}

Complex integrands might be best evaluated by libraries that have no direct connection to QMC software libraries.  In the uncertainty quantification (UQ) space, the recently developed UQ and Model Bridge (UM-Bridge) \cite{UMBridge23}, connects libraries that generate sequences of model parameters, $\bsx_1, \bsx_2, \ldots$ with libraries that evaluate models, $f(\bsx_1), f(\bsx_2), \ldots$, and their derivatives if applicable. UM-Bridge allows coupling of UQ codes to models regardless of the programming languages.

We illustrate in Figure~\ref{CDHJSfigbeam} how UM-Bridge connects QMCPy with a differential equation boundary value problem solver for a cantilevered beam of length~$L$, with a random log effective stiffness $m(\cdot)$. Following ~\cite{ParSee22a}, we assume $m(\cdot)$ is piecewise constant over three regions and the three constants take a $\calU[1,1.2]^3$ distribution.
%, i.e., $m(x) = \left[\begin{matrix} m_1(x), m_2(x), m_3(x) \end{matrix} \right]$ with $m_i(x) \sim \calU[1,1.2]^3$.  
The beam position $x$ is discretized at $N=31$ equidistant points, and
the vertical deflection $u(x)$ when the beam is subject to a given external load  $g(x)$ satisfies the Euler-Bernoulli partial differential equation (PDE):
$$ \frac{\partial^2}{\partial x^2}\left[ \exp(m(x)) \frac{\partial^2  u(x)}{\partial x^2}\right] = g(x),$$
%where $m(x) = \log E(x)$ is the log of an effective stiffness $E(x)$.  The
with the following boundary conditions:
$$ \left.u(x)\right|_{x=0} = 0, \qquad
\left.\frac{\partial u(x)}{\partial x}\right|_{x=0} = 0, \qquad
\left.\frac{\partial^2 u(x)}{\partial x^2}\right|_{x=L} = 0, \qquad  \left.\frac{\partial^3 u(x)}{\partial x^3}\right|_{x=L} = 0. $$
This PDE can be discretized with finite differences yielding a linear system $K(m) \hat{u} = \hat{g}$, where $\hat{u}$ and $\hat{g} \in \mathbb{R}^N$. 
%approximates $u(x)$ and $m(x)$ at the nodes, respectively.
%The forward model $F:\mathbb{R}^3 \rightarrow \mathbb{R}^N $ maps $m = \left[\begin{matrix}m_1, m_2, m_3 \end{matrix}\right]^T$ to $\hat{u}$, while $Q : \mathbb{R}^N \rightarrow \mathbb{R}^2 $ selects the estimated solution at the 10th and 25th nodes, i.e., $ Q(\hat{u}) := \left[\begin{matrix} \hat{u}_{10}, \hat{u}_{25} \end{matrix} \right]^T$. The objective is to determine the distributions of $Q(F(m))$.
The beam model was written in C++ using the MUQ's model framework~\cite{MUQ2}. %It is one of the UM-Bridge's benchmark problem library, available as a container.

@@

Although Python is popular and versatile, it can be slower than highly optimized implementations in languages like Java or C++.  Developers of QMC software in Python would do well to investigate interfaces to optimized code in other languages or consider other code acceleration techniques provided by libraries such as Numba, JAX, Joblib, or Ray.  With time, languages rise and fall in popularity, and those who wish to provide great QMC software may need to pivot to new software environments.
